---
title: "Data Transformation"
author: "Rick Forest"
date: today

execute:
  echo: true
  warning: false
  message: false

format:
  html:
    toc: true
    number-sections: false
    code-fold: true
    code-summary: "Show code"
    code-tools: true
---

**Objective:** Transform the acquired temperature anomaly data into analysis-ready datasets for evaluating global temperature change. This notebook standardizes the NASA GISTEMP data, assigns observations to the REF, CSN, and RCI climate periods, derives temporal and spatial summaries, and saves the transformed datasets used for visualization.

## Initialization

```{r}
#| label: libraries
suppressPackageStartupMessages({
  library(tidyverse)
  library(gt)
  library(fs)
  library(ncdf4)
  library(sf) 
})
```

```{r}
#| label: setup
dir_raw <- path("data", "raw")
dir_transformed <- path("data", "transformed")

file_name_nc <- "gistemp1200_GHCNv4_ERSSTv5.nc"
netcdf_path <- path(dir_raw, file_name_nc)

source("R/climate-periods.R")

analysis_start_year <- min(climate_periods$start_year)
analysis_end_year <- max(climate_periods$end_year)

mollweide_crs <- "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m"

geographic_id_levels <- c("G", "S", "N")

zone_levels <- c(
  "90S-24S", "24S-24N", "24N-90N",
  "90S-64S", "64S-44S", "44S-24S", "24S-EQU",
  "EQU-24N", "24N-44N", "44N-64N", "64N-90N"
)

climate_period_levels <- climate_periods$period_abbr
```

```{r}
#| label: functions
assign_climate_period <- function(year) {
  cut(
    year,
    breaks = c(
      climate_periods$start_year,
      analysis_end_year + 1
    ),
    labels = climate_periods$period_abbr,
    right = FALSE,
    ordered_result = TRUE
  )
}
```

## Import

```{r}
#| label: import-csv
fil <- "GLB.Ts+dSST.csv"
pth <- path(dir_raw, fil)
df_raw_g <- read_csv(pth, skip = 1, na = "***", show_col_types = FALSE)

fil <- "NH.Ts+dSST.csv"
pth <- path(dir_raw, fil)
df_raw_n <- read_csv(pth, skip = 1, na = "***", show_col_types = FALSE)

fil <- "SH.Ts+dSST.csv"
pth <- path(dir_raw, fil)
df_raw_s <- read_csv(pth, skip = 1, na = "***", show_col_types = FALSE)

fil <- "ZonAnn.Ts+dSST.csv"
pth <- path(dir_raw, fil)
df_raw_z <- read_csv(pth, skip = 0, na = "***", show_col_types = FALSE)

```

```{r}
#| label: import-netcdf
nc_data <- ncdf4::nc_open(netcdf_path)
vname <- "tempanomaly"
fillvalue <- ncdf4::ncatt_get(nc_data, vname, "_FillValue")

lon <- ncdf4::ncvar_get(nc_data, "lon")
lat <- ncdf4::ncvar_get(nc_data, "lat")
time <- ncdf4::ncvar_get(nc_data, "time")
values <- ncdf4::ncvar_get(nc_data, vname)

ncdf4::nc_close(nc_data)
```

## Transform

```{r}
#| label: transform-annual

cols_to_keep <- c("Year", "J-D")

fn_annual <- function(df, id) {
  dft <- df |>
    select(all_of(cols_to_keep)) |>
    rename(year = Year) |>
    rename(anomaly = `J-D`) |>
    mutate(geographic_id = id) |>
    mutate(geographic_id = factor(geographic_id, levels = geographic_id_levels)) |>
    mutate(year = as.integer(year)) |>
    mutate(decade = as.integer(year - (year %% 10) + 10)) |>
    mutate(climate_period = assign_climate_period(year)) |>
    select(geographic_id, climate_period, decade, year, anomaly) |>
    drop_na()
}

df1 <- fn_annual(df_raw_g, id = "G")
df2 <- fn_annual(df_raw_n, id = "N")
df3 <- fn_annual(df_raw_s, id = "S")

df_annual <- bind_rows(df1, df2, df3)
df_annual <- df_annual |>
  filter(
    year >= analysis_start_year,
    year <= analysis_end_year
  )
rm(df1, df2, df3)
```

```{r}
#| label: transform-monthly

cols_to_keep <- c(
  "Year",
  "Jan", "Feb", "Mar", "Apr", "May", "Jun",
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)

fn_monthly <- function(df, id) {
  dft <- df |>
    select(all_of(cols_to_keep)) |>
    pivot_longer(
      cols = c(Jan, Feb, Mar, Apr, May, Jun, Jul, Aug, Sep, Oct, Nov, Dec),
      names_to = "month_abbr",
      values_to = "anomaly",
      values_drop_na = TRUE
    ) |>
    rename(year = Year) |>
    mutate(month = match(month_abbr, month.abb)) |>
    mutate(date = as.Date(sprintf("%04d-%02d-01", year, month))) |>
    mutate(geographic_id = id) |>
    mutate(geographic_id = factor(geographic_id, levels = geographic_id_levels)) |>
    mutate(year = as.integer(year)) |>
    mutate(decade = as.integer(year - (year %% 10) + 10)) |>
    mutate(climate_period = assign_climate_period(year)) |>
    select(geographic_id, climate_period, decade, year, month, date, anomaly)
}

df1 <- fn_monthly(df_raw_g, id = "G")
df2 <- fn_monthly(df_raw_n, id = "N")
df3 <- fn_monthly(df_raw_s, id = "S")

df_monthly <- bind_rows(df1, df2, df3)
df_monthly <- df_monthly |>
  filter(
    year >= analysis_start_year,
    year <= analysis_end_year
  )

rm(df1, df2, df3)

```

```{r}
#| label: transform-seasonal

cols_to_keep <- c("Year", "DJF", "MAM", "JJA", "SON")

fn_seasonal <- function(df, id) {
  dft <- df |>
    select(all_of(cols_to_keep)) |>
    pivot_longer(
      cols = c(DJF, MAM, JJA, SON),
      names_to = "months",
      values_to = "anomaly",
      values_drop_na = TRUE
    ) |>
    rename(year = Year) |>
    mutate(geographic_id = id) |>
    mutate(geographic_id = factor(geographic_id, geographic_id_levels)) |>
    mutate(year = as.integer(year)) |>
    mutate(decade = as.integer(year - (year %% 10) + 10)) |>
    mutate(climate_period = assign_climate_period(year)) |>
    mutate(season = case_when(
      geographic_id == "N" & months == "DJF" ~ "Winter",
      geographic_id == "N" & months == "MAM" ~ "Spring",
      geographic_id == "N" & months == "JJA" ~ "Summer",
      geographic_id == "N" & months == "SON" ~ "Autumn",
      geographic_id == "S" & months == "DJF" ~ "Summer",
      geographic_id == "S" & months == "MAM" ~ "Autumn",
      geographic_id == "S" & months == "JJA" ~ "Winter",
      geographic_id == "S" & months == "SON" ~ "Spring"
    )) |>
    select(geographic_id, climate_period, decade, year, season, anomaly)
}

df1 <- fn_seasonal(df_raw_n, id = "N")
df2 <- fn_seasonal(df_raw_s, id = "S")

df_seasonal <- bind_rows(df1, df2)

df_seasonal <- df_seasonal |>
  filter(
    year >= analysis_start_year,
    year <= analysis_end_year
  )

rm(df1, df2)

```

```{r}
#| label: transform-zonal

fn_zonal <- function(df, select_cols, tidy_cols, id) {
  dft <- df |>
    select(all_of(select_cols)) |>
    pivot_longer(
      cols = all_of(tidy_cols),
      names_to = "zone",
      values_to = "anomaly",
      values_drop_na = TRUE
    ) |>
    rename(year = Year) |>
    mutate(zonal_id = id) |>
    mutate(zonal_id = factor(zonal_id)) |>
    mutate(zone = factor(zone, levels = zone_levels)) |>
    mutate(year = as.integer(year)) |>
    mutate(decade = as.integer(year - (year %% 10) + 10)) |>
    mutate(climate_period = assign_climate_period(year)) |>
    select(zonal_id, zone, climate_period, year, anomaly)
}

tidy_cols <- colnames(df_raw_z[, 5:7])
select_cols <- c("Year", tidy_cols)
df_z3 <- fn_zonal(df_raw_z, select_cols, tidy_cols, id = "z3")

tidy_cols <- colnames(df_raw_z[, 8:15])
select_cols <- c("Year", tidy_cols)
df_z8 <- fn_zonal(df_raw_z, select_cols, tidy_cols, id = "z8")

df_zonal <- bind_rows(df_z3, df_z8)
df_zonal <- df_zonal |>
  filter(
    year >= analysis_start_year,
    year <= analysis_end_year
  )
rm(df_z3, df_z8, tidy_cols, select_cols)
```

```{r}
#| label: transform-gridded
values[values == fillvalue$value] <- NA

dates <- as.Date(
  time,
  origin = "1800-01-01"
)

years <- as.integer(
  format(dates, "%Y")
)

netcdf_period <- assign_climate_period(years)

period_levels <- climate_periods$period_abbr

values_period <- simplify2array(
  lapply(
    period_levels,
    function(p) {
      period_index <-
        years >= analysis_start_year &
          years <= analysis_end_year &
          !is.na(netcdf_period) &
          netcdf_period == p

      apply(
        values[
          , ,
          period_index,
          drop = FALSE
        ],
        c(1, 2),
        mean,
        na.rm = TRUE
      )
    }
  )
)

# Grid spacing
lon_resolution <- median(diff(sort(unique(lon))))
lat_resolution <- median(diff(sort(unique(lat))))

# Create one polygon for every netCDF grid cell
grid_geometry <- st_make_grid(
  st_as_sfc(
    st_bbox(
      c(
        xmin = min(lon) - lon_resolution / 2,
        xmax = max(lon) + lon_resolution / 2,
        ymin = min(lat) - lat_resolution / 2,
        ymax = max(lat) + lat_resolution / 2
      ),
      crs = st_crs(4326)
    )
  ),
  cellsize = c(
    lon_resolution,
    lat_resolution
  ),
  what = "polygons"
)

grid_centers <- expand.grid(
  lon = lon,
  lat = lat
)

grid_sf <- st_sf(
  cell_id = seq_len(nrow(grid_centers)),
  grid_centers,
  geometry = grid_geometry,
  crs = 4326
)

df_gridded <- expand.grid(
  cell_id = seq_len(length(lon) * length(lat)),
  climate_period = period_levels
) |>
  mutate(
    avg_anomaly = as.vector(values_period)
  ) |>
  filter(
    is.finite(avg_anomaly)
  ) |>
  mutate(
    climate_period = factor(
      climate_period,
      levels = period_levels,
      ordered = TRUE
    )
  )

sf_gridded <- df_gridded |>
  left_join(
    grid_sf,
    by = "cell_id"
  ) |>
  st_as_sf()

sf_gridded <- sf_gridded |>
  st_transform(mollweide_crs)

sf_gridded <- sf_gridded |>
  mutate(
    area_km2 = as.numeric(
      st_area(geometry)
    ) / 1e6
  )
```

## Aggregate

```{r}
#| label: aggregate-tabular
df_decade <- df_annual |>
  group_by(
    geographic_id,
    decade
  ) |>
  summarize(
    anomaly = mean(
      anomaly,
      na.rm = TRUE
    ),
    n_years = sum(
      !is.na(anomaly)
    ),
    .groups = "drop"
  )
df_hemisphere_period <- df_monthly |>
  filter(
    geographic_id != "G",
    !is.na(climate_period)
  ) |>
  group_by(
    geographic_id,
    climate_period
  ) |>
  summarize(
    avg_anomaly = mean(anomaly, na.rm = TRUE),
    n_months = sum(!is.na(anomaly)),
    .groups = "drop"
  )
df_zone_period <- df_zonal |>
  filter(!is.na(climate_period)) |>
  group_by(
    zonal_id,
    zone,
    climate_period
  ) |>
  summarize(
    avg_anomaly = mean(anomaly, na.rm = TRUE),
    n_years = sum(!is.na(anomaly)),
    .groups = "drop"
  )

df_hemisphere_period <- df_hemisphere_period |>
  mutate(
    climate_period = factor(
      climate_period,
      levels = climate_period_levels,
      ordered = TRUE
    )
  )
df_zone_period <- df_zone_period |>
  mutate(
    climate_period = factor(
      climate_period,
      levels = climate_period_levels,
      ordered = TRUE
    )
  )
```

## Save

```{r}
#| label: save
write_rds(df_monthly, path(dir_transformed, "monthly.rds"))
write_rds(df_annual, path(dir_transformed, "annual.rds"))
write_rds(df_decade, path(dir_transformed, "decade.rds"))
write_rds(df_seasonal, path(dir_transformed, "seasonal.rds"))
write_rds(df_hemisphere_period, path(dir_transformed, "hemisphere.rds"))
write_rds(df_zone_period, path(dir_transformed, "zone.rds"))
write_rds(sf_gridded, path(dir_transformed, "gridded.rds"))
```

## Transformation Summary

```{r}
#| label: table-descriptions

transformed_descriptions <- tribble(
  ~filename,                     ~dataset,
  "monthly.rds",                 "Monthly temperature anomalies",
  "annual.rds",                  "Annual temperature anomalies",
  "decade.rds",                  "Decadal temperature anomalies",
  "seasonal.rds",                "Seasonal temperature anomalies",
  "hemisphere.rds",              "Hemisphere anomalies by climate period",
  "zone.rds",                    "Latitude-zone anomalies by climate period",
  "gridded.rds",                 "Gridded anomalies by climate period"
)

transformed_catalog <- tibble(
  path = dir_ls(
    dir_transformed,
    type = "file"
  )
)

file_metadata <- file_info(transformed_catalog$path)

transformed_catalog <- transformed_catalog |>
  mutate(
    filename = path_file(path),
    format = path_ext(filename),
    directory = path_dir(path),
    transformation_datetime = format(
      file_metadata$modification_time,
      "%Y-%m-%d %H:%M"
    ),
    file_size_kb = round(
      as.numeric(file_metadata$size) / 1024,
      0
    )
  ) |>
  left_join(
    transformed_descriptions,
    by = "filename"
  ) |>
  arrange(filename) |>
  select(
    dataset,
    filename,
    format,
    directory,
    transformation_datetime,
    file_size_kb
  )

```

```{r}
#| label: transformed-data-summary
#| echo: false

transformed_catalog |>
  gt() |>
  cols_label(
    dataset = "Dataset",
    filename = "Filename",
    format = "File Format",
    directory = "Local Directory",
    transformation_datetime = "Transformed",
    file_size_kb = "Size (KB)"
  ) |>
  fmt_number(
    columns = file_size_kb,
    decimals = 0
  ) |>
  cols_align(
    align = "center",
    columns = c(
      format,
      transformation_datetime,
      file_size_kb
    )
  ) |>
  tab_header(
    title = "Transformed Data Files",
    subtitle = "Files produced by the data transformation workflow"
  )
```
