---
title: "Data Transformation"
author: "Rick Forest"
date: today
format:
  html:
    toc: true
    number-sections: false
    code-fold: true
    code-tools: true
---

This step transforms the acquired temperature anomaly data.

## Initialization

### Libraries

```{r}
#| label: libraries
suppressPackageStartupMessages({
  library(tidyverse) # package collection used in transforming tabular data
  library(ncdf4)     # process netCDF files
  library(gt)        # package used to create and display formatted tables
  library(sf)        # package used in transforming spatial data
})
```

### Directories

```{r}
#| label: directories
dir_raw <- file.path("data", "raw")
dir_transformed <- file.path("data", "transformed")
```

### Paths

```{r}
#| label: paths
file_name_nc = "gistemp1200_GHCNv4_ERSSTv5.nc"
netCDF_path <- file.path(dir_raw, file_name_nc)
```

### Climate Periods

```{r}
#| label: R-climate-periods
source("R/climate-periods.R")
```

### Analysis Start Year

```{r}
#| label: analysis-start-year
analysis_start_year <- min(climate_periods$start_year)
```

### Coordinate Reference System

```{r}
#| label: coordinate-reference-system
mollweide_crs <- "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m" 
```

### Functions

```{r}
#| label: function-assign-climate-period
assign_climate_period <- function(year) {
  cut(
    year,
    breaks = c(
      climate_periods$start_year,
      Inf
    ),
    labels = climate_periods$climate_period,
    right = FALSE,
    ordered_result = TRUE
  )
}
```

### Factors

```{r}
#| label: factors
geographic_id_levels <- c("G","S","N")

zone_levels <- c( "90S-24S", "24S-24N", "24N-90N",
                  "90S-64S", "64S-44S", "44S-24S", "24S-EQU",
                  "EQU-24N", "24N-44N", "44N-64N", "64N-90N")

climate_period_levels <- c("1", "2", "3", "4", "5")
```

## Import

### Import csv

```{r}
#| label: import-csv
fil <- "GLB.Ts+dSST.csv"
pth <- file.path(dir_raw, fil)
df_raw_g <- read_csv(pth, skip=1, na = "***", show_col_types = FALSE)
names(df_raw_g)

fil <- "NH.Ts+dSST.csv"
pth <- file.path(dir_raw, fil)
df_raw_n  <- read_csv(pth, skip=1, na = "***", show_col_types = FALSE)
names(df_raw_n)

fil <- "SH.Ts+dSST.csv"
pth <- file.path(dir_raw, fil)
df_raw_s  <- read_csv(pth, skip=1, na = "***", show_col_types = FALSE)
names(df_raw_s)

fil <- "ZonAnn.Ts+dSST.csv"
pth <- file.path(dir_raw, fil)
df_raw_z <- read_csv(pth, skip=0, na = "***", show_col_types = FALSE)
names(df_raw_z)
```

### Import netCDF

#### Open netCDF

```{r}
#| label: netCDF-import
nc_data <- ncdf4::nc_open(netCDF_path)
vname <- "tempanomaly"
fillvalue <- ncdf4::ncatt_get(nc_data,vname,"_FillValue")
```

#### Extract netCDF variables

```{r}
#| label: netCDF-variables
lon <- ncdf4::ncvar_get(nc_data,"lon")
lat <- ncdf4::ncvar_get(nc_data,"lat")
time <- ncdf4::ncvar_get(nc_data,"time")
values <- ncdf4::ncvar_get(nc_data, vname)
```

#### Close netCDF

```{r}
#| label: netCDF-close
ncdf4::nc_close(nc_data)
```

## Transform

### Transform tabular

#### Global and Hemisphere Annual

```{r}
#| label: tabular-annual

cols_to_keep <- c("Year", "J-D")

fn_annual <- function(df, id) {
  dft <- df |> 
    select(all_of(cols_to_keep)) |>
    rename(year = Year) |>
    rename(anomaly = `J-D`) |>
    mutate(geographic_id = id) |> 
    mutate(geographic_id = factor(geographic_id, levels=geographic_id_levels)) |> 
    mutate(year = as.integer(year)) |>
    mutate(decade = as.integer(year - (year %% 10) + 10)) |>
    mutate(climate_period = assign_climate_period(year)) |>
    select(geographic_id, climate_period, decade, year, anomaly) |> 
    drop_na()
  return(dft)
}

df1 <- fn_annual(df_raw_g, id="G")
df2 <- fn_annual(df_raw_n, id="N")
df3 <- fn_annual(df_raw_s, id="S")

df_annual <- bind_rows(df1, df2, df3)
df_annual <- df_annual |> filter(year >= analysis_start_year)
rm(df1, df2, df3)
```

#### Global and Hemisphere Monthly

```{r}
#| label: tabular-monthly

cols_to_keep <- c("Year",
  "Jan", "Feb", "Mar", "Apr", "May", "Jun",
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

fn_monthly <- function(df, id) {
  dft <- df |> 
    select(all_of(cols_to_keep)) |>
    pivot_longer(
      cols=c(Jan,Feb,Mar,Apr,May,Jun,Jul,Aug,Sep,Oct,Nov,Dec), 
      names_to = "month_abbr",
      values_to = "anomaly",
      values_drop_na = TRUE) |>
    rename(year = Year) |>
    mutate(month = match(month_abbr,month.abb)) |> 
    mutate(date = as.Date(sprintf("%04d-%02d-01",year,month))) |> 
    mutate(geographic_id = id) |> 
    mutate(geographic_id = factor(geographic_id, levels=geographic_id_levels)) |> 
    mutate(year = as.integer(year)) |>
    mutate(decade = as.integer(year - (year %% 10) + 10)) |>
    mutate(climate_period = assign_climate_period(year)) |>
    select(geographic_id, climate_period, decade, year, month, date, anomaly)

  return(dft)
}

df1 <- fn_monthly(df_raw_g, id="G")
df2 <- fn_monthly(df_raw_n, id="N")
df3 <- fn_monthly(df_raw_s, id="S")

df_monthly <- bind_rows(df1, df2, df3)
df_monthly <- df_monthly |> filter(year >= analysis_start_year)
rm(df1, df2, df3)
```

#### Seasonal Annual

Meteorological seasons are hemisphere dependent:  

  - NH winter, SH summer: December, January, February. 
  - NH spring, SH autumn: March, April, May. 
  - NH summer, SH winter: June, July, August. 
  - NH autumn, SH spring: September, October, November. 

```{r}
#| label: tabular-seasonal

cols_to_keep <- c("Year", "DJF", "MAM", "JJA", "SON")

fn_seasonal <- function(df, id) {
  dft <- df |> 
    select(all_of(cols_to_keep)) |>
    pivot_longer(
      cols=c(DJF, MAM, JJA, SON), 
      names_to = "months",
      values_to = "anomaly",
      values_drop_na = TRUE) |>
    rename(year = Year) |>
    mutate(geographic_id = id) |> 
    mutate(geographic_id = factor(geographic_id, geographic_id_levels)) |> 
    mutate(year = as.integer(year)) |>
    mutate(decade = as.integer(year - (year %% 10) + 10)) |>
    mutate(climate_period = assign_climate_period(year)) |>
    mutate(season = case_when(
      geographic_id == "N" & months == "DJF" ~ "Winter",
      geographic_id == "N" & months == "MAM" ~ "Spring",
      geographic_id == "N" & months == "JJA" ~ "Summer",
      geographic_id == "N" & months == "SON" ~ "Autumn",
      geographic_id == "S" & months == "DJF" ~ "Summer",
      geographic_id == "S" & months == "MAM" ~ "Autumn",
      geographic_id == "S" & months == "JJA" ~ "Winter",
      geographic_id == "S" & months == "SON" ~ "Spring"
    ))|>
    select(geographic_id, climate_period, decade, year, season, anomaly)
  return(dft)
}

df1 <- fn_seasonal(df_raw_n, id="N")
df2 <- fn_seasonal(df_raw_s, id="S")

df_seasonal <- bind_rows(df1, df2)

df_seasonal <- df_seasonal |> filter(year >= analysis_start_year)
rm(df1, df2)
```

#### Zonal Annual 

```{r}
#| label: tabular-zonal

fn_zonal <- function(df,select_cols,tidy_cols,id) {
  dft <- df |> 
    select(all_of(select_cols)) |>
    pivot_longer(
      cols =  all_of(tidy_cols),
      names_to = "zone",
      values_to = "anomaly",
      values_drop_na = TRUE) |> 
    rename(year = Year) |>
    mutate(zonal_id = id) |>
    mutate(zonal_id = factor(zonal_id)) |>
    mutate(zone = factor(zone, levels=zone_levels)) |>
    mutate(year = as.integer(year)) |>
    mutate(decade = as.integer(year - (year %% 10) + 10)) |> 
    mutate(climate_period = assign_climate_period(year)) |>
    select(zonal_id, zone, climate_period, year, anomaly)
  return(dft) 
}

tidy_cols <- colnames(df_raw_z[,5:7])
select_cols <- c("Year",tidy_cols)
df_z3 <- fn_zonal(df_raw_z, select_cols, tidy_cols, id = "z3")

tidy_cols <- colnames(df_raw_z[,8:15])
select_cols <- c("Year",tidy_cols)
df_z8 <- fn_zonal(df_raw_z, select_cols, tidy_cols, id = "z8")

df_zonal <- bind_rows(df_z3, df_z8)
df_zonal <- df_zonal |> filter(year >= analysis_start_year)
rm(df_z3, df_z8, tidy_cols, select_cols)
```

### Transform gridded

#### Identify missing values

```{r}
#| label: gridded-missing-values
values[values==fillvalue$value] <- NA
```

#### Transform dates

```{r}
#| label: gridded-time

dates <- as.Date(
  time,
  origin = "1800-01-01"
)

years <- as.integer(
  format(dates, "%Y")
)
```

#### Assign climate period

```{r}
#| label: gridded-assign-period

netCDF_period <- assign_climate_period(years)
```

#### Aggregate to climate period

```{r}
#| label: gridded-aggregate-climate-period

period_levels <- climate_periods$climate_period

values_period <- simplify2array(
  lapply(
    period_levels,
    function(p) {

      period_index <-
        years >= analysis_start_year &
        !is.na(netCDF_period) &
        netCDF_period == p

      apply(
        values[
          ,
          ,
          period_index,
          drop = FALSE
        ],
        c(1, 2),
        mean,
        na.rm = TRUE
      )
    }
  )
)
```

#### Convert to spatial data frame

##### Create grid polygons

```{r}
#| label: create-grid-polygons

# Grid spacing
lon_resolution <- median(diff(sort(unique(lon))))
lat_resolution <- median(diff(sort(unique(lat))))

# Create one polygon for every netCDF grid cell
grid_geometry <- st_make_grid(
  st_as_sfc(
    st_bbox(
      c(
        xmin = min(lon) - lon_resolution / 2,
        xmax = max(lon) + lon_resolution / 2,
        ymin = min(lat) - lat_resolution / 2,
        ymax = max(lat) + lat_resolution / 2
      ),
      crs = st_crs(4326)
    )
  ),
  cellsize = c(
    lon_resolution,
    lat_resolution
  ),
  what = "polygons"
)

grid_centers <- expand.grid(
  lon = lon,
  lat = lat
)

grid_sf <- st_sf(
  cell_id = seq_len(nrow(grid_centers)),
  grid_centers,
  geometry = grid_geometry,
  crs = 4326
)
```

##### Attach climate period anomalies

```{r}
#| label: expand-grid
df_gridded <- expand.grid(
  cell_id = seq_len(length(lon) * length(lat)),
  climate_period = period_levels
) |>
  mutate(
    avg_anomaly = as.vector(values_period)
  ) |>
  filter(
    is.finite(avg_anomaly)
  ) |>
  mutate(
    climate_period = factor(
      climate_period,
      levels = period_levels,
      ordered = TRUE
    )
  )

sf_gridded <- df_gridded |>
  left_join(
    grid_sf,
    by = "cell_id"
  ) |>
  st_as_sf()
```

##### Equal Area Projection

```{r}
#| label: gridded-equal-area-projection
sf_gridded <- sf_gridded |>
   st_transform(mollweide_crs)
```

##### Calculate cell areas

```{r}
#| label: area-km
sf_gridded <- sf_gridded |>
  mutate(
    area_km2 = as.numeric(
      st_area(geometry)
    ) / 1e6
  )
```

## Aggregate

### Aggregate tabular

```{r}
#| label: aggregate-tabular-data
df_decade <- df_annual |>
  group_by(
    geographic_id,
    decade
  ) |>
  summarize(
    anomaly = mean(
      anomaly,
      na.rm = TRUE
    ),
    n_years = sum(
      !is.na(anomaly)
    ),
    .groups = "drop"
  )
df_hemisphere_period <- df_monthly |>
  filter(geographic_id != "G",
    !is.na(climate_period)) |>
  group_by(
    geographic_id,
    climate_period
  ) |>
  summarize(
    avg_anomaly = mean(anomaly, na.rm = TRUE),
    n_months = sum(!is.na(anomaly)),
    .groups = "drop"
  )
df_zone_period <- df_zonal |>
  filter(!is.na(climate_period)) |>
  group_by(
    zonal_id,
    zone,
    climate_period
  ) |>
  summarize(
    avg_anomaly = mean(anomaly, na.rm = TRUE),
    n_years = sum(!is.na(anomaly)),
    .groups = "drop"
  )
```

### Aggregate climate periods

```{r}
#| label: aggregate-tabular-climate-periods
df_hemisphere_period <- df_hemisphere_period |>
  mutate(
    climate_period = factor(
      climate_period,
      levels = climate_period_levels,
      ordered = TRUE
    )
  )
df_zone_period <- df_zone_period |>
  mutate(
    climate_period = factor(
      climate_period,
      levels = climate_period_levels,
      ordered = TRUE
    )
  )
```

## Save

```{r}
#| label: save
# saving as rds instead of csv preserves the factor designations
write_rds(df_monthly, file.path(dir_transformed, "monthly.rds"))
write_rds(df_annual, file.path(dir_transformed, "annual.rds"))
write_rds(df_decade, file.path(dir_transformed, "decade.rds"))
write_rds(df_seasonal, file.path(dir_transformed, "seasonal.rds"))
write_rds(df_hemisphere_period, file.path(dir_transformed,"hemisphere.rds"))
write_rds(df_zone_period, file.path(dir_transformed,"zone.rds"))
write_rds(sf_gridded, file.path(dir_transformed,"gridded.rds"))
```

## Transformed Data Catalog

### Lookup table

```{r}
#| label: table-descriptions
transformed_descriptions <- tribble(
  ~filename,                     ~dataset,
  "monthly.rds",                 "Monthly temperature anomalies",
  "annual.rds",                  "Annual temperature anomalies",
  "decade.rds",                  "Decadal temperature anomalies",
  "seasonal.rds",                "Seasonal temperature anomalies",
  "hemisphere.rds",              "Hemisphere anomalies by climate period",
  "zone.rds",                    "Latitude-zone anomalies by climate period",
  "gridded.rds",                 "Gridded anomalies by climate period"
)
```

### Create catalog

```{r}
#| label: transformed-data-catalog

transformed_catalog <- tibble(
  path = list.files(
    dir_transformed,
    full.names = TRUE
  )
) |>
  filter(
    !file.info(path)$isdir
  ) |>
  mutate(
    filename = basename(path),
    format = tools::file_ext(filename),
    directory = dirname(path),
    transformation_datetime = format(
      file.info(path)$mtime,
      "%Y-%m-%d %H:%M"
    ),
    file_size_kb = round(
      file.info(path)$size / 1024,
      0
    )
  ) |>
  left_join(
    transformed_descriptions,
    by = "filename"
  ) |>
  arrange(filename) |>
  select(
    dataset,
    filename,
    format,
    directory,
    transformation_datetime,
    file_size_kb
  )
```

```{r}
#| label: transformed-data-summary

transformed_catalog |>
  gt() |>
  cols_label(
    dataset = "Dataset",
    filename = "Filename",
    format = "File Format",
    directory = "Local Directory",
    transformation_datetime = "Transformed",
    file_size_kb = "Size (KB)"
  ) |>
  fmt_number(
    columns = file_size_kb,
    decimals = 0
  ) |>
  cols_align(
    align = "center",
    columns = c(
      format,
      transformation_datetime,
      file_size_kb
    )
  ) |>
  tab_header(
    title = "Transformed Data Files",
    subtitle = "Files produced by the data transformation workflow"
  )
```