---
title: "Data Visualization"
author: "Rick Forest"
date: today

execute:
  echo: true
  warning: false
  message: false

format:
  html:
    toc: true
    number-sections: false
    code-fold: true
    code-summary: "Show code"
    code-tools: true
---
---

**Objective:** Visualize the transformed temperature anomaly data to examine how global temperature change varies over time, across climate periods, and by geographic and seasonal characteristics. This notebook summarizes temporal trends, temperature distributions, hemispheric and latitude-zone differences, seasonal patterns, affected surface area, and concludes with global maps comparing the REF, CSN, and RCI climate periods.

## Initialization

```{r}
#| label: libraries
suppressPackageStartupMessages({
  library(tidyverse)
  library(gt)
  library(fs)
  library(patchwork) 
  library(pals) 
  library(viridisLite) 
  library(sf)
})
```


```{r}
#| label: setup
build_maps <- TRUE

dir_raw <- path("data", "raw")
dir_transformed <- path("data", "transformed")

source("R/climate-periods.R")

theme_climate <- theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 20, face = "bold", margin = margin(b = 6)),
    plot.subtitle = element_text(size = 16, margin = margin(b = 10)),
    axis.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 12, color = "black"),
    axis.ticks = element_line(linewidth = 0.5, color = "black"),
    axis.ticks.length = grid::unit(4, "pt"),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    legend.key.width = unit(1.2, "cm"),
    legend.key.height = unit(0.5, "cm"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )

legend_climate <- guide_colorbar(
  direction = "horizontal",
  title.position = "top",
  title.hjust = 0.5,
  barwidth = unit(10, "cm"),
  barheight = unit(0.45, "cm"),
  ticks = TRUE
)

# Diverging temperature-anomaly palette
anomaly_colors <- pals::coolwarm(11)

# Sequential climate-period palette
climate_period_palette <- setNames(
  viridisLite::viridis(
    nrow(climate_periods),
    option = "D",
    begin = 0.25,
    end = 0.80
  ),
  climate_periods$period_abbr
)

# Categorical hemisphere palette
hemisphere_palette <- c(
  "N" = viridis(5)[2],
  "S" = viridis(5)[4]
)

# Categorical season palette
season_palette <- c(
  "Winter" = "#0072B2", # blue
  "Spring" = "#009E73", # bluish green
  "Summer" = "#E69F00", # orange
  "Autumn" = "#CC79A7" # reddish purple
)

anomaly_fill_scale <- function(limits) {
  scale_fill_gradient2(
    low = anomaly_colors[1],
    mid = anomaly_colors[6],
    high = anomaly_colors[11],
    midpoint = 0,
    limits = limits,
    name = "Anomaly (°C)",
    guide = legend_climate
  )
}

mollweide_crs <- "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m"

zone_levels_3 <- c(
  "24N-90N",
  "24S-24N",
  "90S-24S"
)
zone_levels_8 <- c(
  "64N-90N",
  "44N-64N",
  "24N-44N",
  "EQU-24N",
  "24S-EQU",
  "44S-24S",
  "64S-44S",
  "90S-64S"
)

zone_labels_3 <- c(
  "24N-90N" = "24°N–90°N",
  "24S-24N" = "24°S–24°N",
  "90S-24S" = "90°S–24°S"
)

zone_labels_8 <- c(
  "64N-90N" = "64°N–90°N",
  "44N-64N" = "44°N–64°N",
  "24N-44N" = "24°N–44°N",
  "EQU-24N" = "Equator–24°N",
  "24S-EQU" = "24°S–Equator",
  "44S-24S" = "44°S–24°S",
  "64S-44S" = "64°S–44°S",
  "90S-64S" = "90°S–64°S"
)
```

## Import

```{r}
#| label: import
df_monthly <- read_rds(path(dir_transformed, "monthly.rds"))
df_annual <- read_rds(path(dir_transformed, "annual.rds"))
df_decade <- read_rds(path(dir_transformed, "decade.rds"))
df_seasonal <- read_rds(path(dir_transformed, "seasonal.rds"))
df_hemisphere <- read_rds(path(dir_transformed, "hemisphere.rds"))
df_zone <- read_rds(path(dir_transformed, "zone.rds"))
sf_gridded <- read_rds(path(dir_transformed, "gridded.rds"))
sf_coastline <- sf::read_sf(path(dir_raw, "global-coastline.gpkg"))

latest_year <- max(df_monthly$year, na.rm = TRUE)
```

```{r}
#| label: climate-periods-labels

climate_period_labels <- climate_periods |>
  arrange(start_year) |>
  transmute(
    period_abbr,
    label = paste0(
      period_label,
      " (",
      start_year,
      "–",
      end_year,
      ")"
    )
  ) |>
  tibble::deframe()
```

```{r}
#| label: filter
df_glb_mth <- df_monthly |> filter(geographic_id == "G")
df_glb_ann <- df_annual |> filter(geographic_id == "G")
df_glb_dec <- df_decade |> filter(geographic_id == "G")
df_glb_period <- df_monthly |> filter(geographic_id == "G", !is.na(climate_period))
```

## Visualizations

### Global Time Series

### Global Time Series

Monthly, annual, and decadal temperature anomalies show how the global temperature signal has evolved over the analysis period.

```{r}
#| label: v1-color-limits
glb_color_limits <- range(
  df_glb_mth$anomaly,
  df_glb_ann$anomaly,
  df_glb_dec$anomaly,
  na.rm = TRUE
)
```

```{r}
#| label: v1-color-scale
glb_color_scale <- scale_color_gradientn(
  colours = c(
    "#3B4CC0", # strong blue
    "#5B74C9", # blue
    "#8FA4D7", # light blue
    "#F7F7F7", # neutral at zero
    "#D98D8D", # light red
    "#C44747", # red
    "#B40426" # strong red
  ),
  values = scales::rescale(
    c(
      glb_color_limits[1],
      -0.10,
      -0.03,
      0,
      0.03,
      0.10,
      glb_color_limits[2]
    ),
    from = glb_color_limits
  ),
  limits = glb_color_limits,
  name = "Anomaly (°C)",
  guide = legend_climate
)
```

```{r}
#| label: v1-axis-limits
mth_range <- range(df_glb_mth$anomaly, na.rm = TRUE)
mth_buffer <- 0.05 * diff(mth_range)
mth_y_limits <- mth_range + c(-mth_buffer, mth_buffer)
ann_dec_range <- range(df_glb_ann$anomaly, df_glb_dec$anomaly, na.rm = TRUE)
ann_dec_buffer <- 0.05 * diff(ann_dec_range)
ann_dec_y_limits <- ann_dec_range + c(-ann_dec_buffer, ann_dec_buffer)
```

```{r}
#| label: v1-create-statistics

v1_statistics <- tibble(
  statistic = c(
    "Average",
    "Maximum",
    "Minimum"
  ),
  monthly = c(
    mean(df_glb_mth$anomaly, na.rm = TRUE),
    max(df_glb_mth$anomaly, na.rm = TRUE),
    min(df_glb_mth$anomaly, na.rm = TRUE)
  ),
  annual = c(
    mean(df_glb_ann$anomaly, na.rm = TRUE),
    max(df_glb_ann$anomaly, na.rm = TRUE),
    min(df_glb_ann$anomaly, na.rm = TRUE)
  ),
  decadal = c(
    mean(df_glb_dec$anomaly, na.rm = TRUE),
    max(df_glb_dec$anomaly, na.rm = TRUE),
    min(df_glb_dec$anomaly, na.rm = TRUE)
  )
)
```

```{r}
#| label: v1-create-table
v1_summary <- v1_statistics |>
  gt() |>
  cols_label(
    statistic = "Statistic",
    monthly = "Monthly",
    annual = "Annual",
    decadal = "Decadal"
  ) |>
  fmt_number(
    columns = c(monthly, annual, decadal),
    decimals = 2
  ) |>
  cols_align(
    align = "left",
    columns = statistic
  ) |>
  cols_align(
    align = "center",
    columns = c(monthly, annual, decadal)
  ) |>
  cols_width(
    statistic ~ pct(40),
    monthly ~ pct(20),
    annual ~ pct(20),
    decadal ~ pct(20)
  ) |>
  tab_header(
    title = "Global Temperature Summary"
  ) |>
  tab_source_note(
    source_note = "Values are temperature anomalies in °C."
  ) |>
  tab_options(
    table.width = pct(100),
    table.font.size = px(16)
  )
```

```{r}
#| label: v1-create-monthly-plot
plt_glb_mth <- df_glb_mth |>
  ggplot(aes(x = date, y = anomaly)) +
  geom_point(
    aes(color = anomaly),
    size = 1.3,
    alpha = 0.8
  ) +
  geom_smooth(
    method = "loess",
    span = 0.75,
    color = "grey20",
    linewidth = 1.1,
    se = FALSE
  ) +
  glb_color_scale +
  coord_cartesian(ylim = mth_y_limits) +
  labs(
    subtitle = "Monthly",
    x = "Year",
    y = "Anomaly (°C)"
  ) +
  theme_climate
```

```{r}
#| label: v1-create-annual-plot
plt_glb_ann <- df_glb_ann |>
  ggplot(aes(x = year, y = anomaly)) +
  geom_line(
    color = "grey20",
    linewidth = 0.7,
    alpha = 0.7
  ) +
  geom_point(
    aes(color = anomaly),
    size = 2.0
  ) +
  glb_color_scale +
  coord_cartesian(ylim = ann_dec_y_limits) +
  labs(
    subtitle = "Annual",
    x = "Year",
    y = "Anomaly (°C)"
  ) +
  theme_climate
```

```{r}
#| label: v1-create-decadal-plot
plt_glb_dec <- df_glb_dec |>
  ggplot(aes(x = decade, y = anomaly)) +
  geom_line(
    color = "grey20",
    linewidth = 1.1,
    alpha = 0.7
  ) +
  geom_point(
    aes(color = anomaly),
    size = 3.0
  ) +
  glb_color_scale +
  coord_cartesian(ylim = ann_dec_y_limits) +
  labs(
    subtitle = "Decadal",
    x = "Decade",
    y = NULL
  ) +
  theme_climate +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

```{r}
#| label: v1-create-composition
#| echo: false

plot_design_1 <- "
AAAAAA
AAAAAA
AAAAAA
BBBCCC
BBBCCC
"

v1_plot_composition <-
  plt_glb_mth +
  plt_glb_ann +
  plt_glb_dec +
  plot_layout(
    design = plot_design_1,
    guides = "collect"
  ) +
  plot_annotation(
    title = "Global Temperature Anomalies",
    subtitle = "Monthly, annual, and decadal temperature anomalies",
    theme = theme_climate
  )
```

```{r}
#| label: v1-display-table
#| echo: false
v1_summary
```

```{r}
#| label: v1-display-composition
#| fig-width: 14
#| fig-height: 12
#| echo: false
v1_plot_composition
```

### Distribution

Temperature anomaly distributions compare how the location and spread of observed anomalies differ among the three climate periods.

```{r}
#| label: v2-create-statistics
v2_statistics <- df_glb_mth |>
  group_by(climate_period) |>
  summarize(
    median = median(anomaly, na.rm = TRUE),
    q1 = quantile(anomaly, 0.25, na.rm = TRUE),
    q3 = quantile(anomaly, 0.75, na.rm = TRUE),
    iqr = IQR(anomaly, na.rm = TRUE),
    lower_fence = q1 - 1.5 * iqr,
    upper_fence = q3 + 1.5 * iqr,
    outliers = sum(
      anomaly < lower_fence |
        anomaly > upper_fence,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

v2_statistics_table <- v2_statistics |>
  select(
    climate_period,
    median,
    q1,
    q3,
    iqr,
    outliers
  ) |>
  pivot_longer(
    cols = -climate_period,
    names_to = "statistic",
    values_to = "value"
  ) |>
  mutate(
    statistic = recode(
      statistic,
      median = "Median",
      q1 = "Q1 (25th percentile)",
      q3 = "Q3 (75th percentile)",
      iqr = "Interquartile range",
      outliers = "Outliers"
    ),
    climate_period = climate_period_labels[
      as.character(climate_period)
    ]
  ) |>
  pivot_wider(
    names_from = climate_period,
    values_from = value
  )
```

```{r}
#| label: v2-create-table
v2_summary <- v2_statistics_table |>
  gt() |>
  cols_label(
    statistic = "Statistic"
  ) |>
  fmt_number(
    rows = !statistic %in% c("Outliers"),
    columns = -statistic,
    decimals = 2
  ) |>
  fmt_number(
    rows = statistic == "Outliers",
    columns = -statistic,
    decimals = 0
  ) |>
  cols_align(
    align = "left",
    columns = statistic
  ) |>
  cols_align(
    align = "center",
    columns = -statistic
  ) |>
  tab_header(
    title = "Distribution Summary by Climate Period"
  ) |>
  tab_source_note(
    source_note = "Temperature anomaly statistics are in °C; outliers are counts."
  )
```

```{r}
#| label: v2-create-boxplot
plt_period_box <- df_glb_period |>
  ggplot(
    aes(
      x = climate_period,
      y = anomaly,
      fill = climate_period
    )
  ) +
  geom_hline(
    yintercept = 0,
    color = "grey45",
    linewidth = 0.5,
    linetype = "dashed"
  ) +
  geom_boxplot(
    alpha = 0.8,
    width = 0.65,
    outlier.alpha = 0.45,
    show.legend = FALSE
  ) +
  scale_fill_manual(
    values = climate_period_palette,
    labels = climate_period_labels,
    name = "Climate Period"
  ) +
  scale_x_discrete(
    labels = climate_period_labels
  ) +
  labs(
    x = "Climate Period",
    y = "Anomaly (°C)"
  )
v2_display_boxplot <-
  plt_period_box +
    plot_annotation(
      title = "Distribution summaries by climate period",
      subtitle = "Median, interquartile range, and outliers"
    ) &
    theme_climate
```

```{r}
#| label: v2-create-histogram
plt_period_hist <- df_glb_period |>
  ggplot(
    aes(
      x = anomaly,
      fill = climate_period
    )
  ) +
  geom_histogram(
    binwidth = 0.1,
    boundary = 0,
    color = "white",
    linewidth = 0.2,
    show.legend = FALSE
  ) +
  geom_vline(
    xintercept = 0,
    color = "grey45",
    linewidth = 0.5,
    linetype = "dashed"
  ) +
  facet_wrap(
    ~climate_period,
    ncol = 1,
    labeller = as_labeller(climate_period_labels),
    scales = "free_y"
  ) +
  scale_fill_manual(
    values = climate_period_palette,
    labels = climate_period_labels,
    name = "Climate Period"
  ) +
  labs(
    x = "Anomaly (°C)",
    y = "Monthly observations"
  ) +
  theme(
    strip.text = element_text(
      size = 13,
      face = "bold"
    )
  )
v2_display_histogram <-
  plt_period_hist +
    plot_annotation(
      title = "Distribution shapes by climate period",
      subtitle = "Frequency distributions of monthly temperature anomalies"
    ) &
    theme_climate
```

```{r}
#| label: v2-display-summary
#| echo: false
v2_summary
```

```{r}
#| label: v2-display-boxplot
#| fig-width: 12
#| fig-height: 8
#| echo: false
v2_display_boxplot
```

```{r}
#| label: v2-display-histogram
#| fig-width: 12
#| fig-height: 10
#| echo: false
v2_display_histogram
```

### Hemispheres

Hemisphere comparisons show whether warming patterns differ between the Northern and Southern Hemispheres.

```{r}
#| label: v3-create-statistics

v3_statistics <- df_hemisphere |>
  mutate(
    hemisphere = recode(
      geographic_id,
      "N" = "Northern",
      "S" = "Southern"
    ),
    hemisphere = factor(
      hemisphere,
      levels = c(
        "Northern",
        "Southern"
      )
    ),
    climate_period =
      climate_period_labels[
        as.character(climate_period)
      ]
  ) |>
  select(
    hemisphere,
    climate_period,
    avg_anomaly
  ) |>
  pivot_wider(
    names_from = climate_period,
    values_from = avg_anomaly
  )
```

```{r}
#| label: v3-create-table
v3_summary <- v3_statistics |>
  gt() |>
  cols_label(
    hemisphere = "Hemisphere"
  ) |>
  fmt_number(
    columns = -hemisphere,
    decimals = 2
  ) |>
  cols_align(
    align = "left",
    columns = hemisphere
  ) |>
  cols_align(
    align = "center",
    columns = -hemisphere
  ) |>
  tab_header(
    title = "Average Hemisphere Temperature Anomaly"
  ) |>
  tab_source_note(
    source_note =
      "Values are average temperature anomalies (°C) relative to the NASA GISTEMP 1951–1980 baseline."
  )
```

```{r}
#| label: v3-create-visualizations

df_hemisphere <- df_hemisphere |>
  mutate(geographic_id = factor(geographic_id, levels = c("S", "N")))

# Consistent vertical-axis limits for both plots
df_hemisphere_avg <- df_hemisphere |>
  group_by(geographic_id) |>
  summarize(avg_anomaly = mean(avg_anomaly, na.rm = TRUE), .groups = "drop")
y_limits <- range(
  c(
    0,
    df_hemisphere_avg$avg_anomaly,
    df_hemisphere$avg_anomaly
  ),
  na.rm = TRUE
)

# Add space above and below the bars
y_expansion <- expansion(mult = c(0.05, 0.10))

# Hemisphere average
plt_hemisphere_avg <- df_hemisphere_avg |>
  ggplot(aes(
    x = geographic_id,
    y = avg_anomaly,
    fill = geographic_id
  )) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = hemisphere_palette) +
  scale_y_continuous(
    limits = y_limits,
    expand = y_expansion
  ) +
  labs(
    x = "Hemisphere",
    y = "Anomaly (°C)"
  ) +
  theme_climate

plt_hemisphere_period <- df_hemisphere |>
  ggplot(
    aes(
      x = climate_period,
      y = avg_anomaly,
      fill = geographic_id
    )
  ) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7
  ) +
  scale_fill_manual(
    values = hemisphere_palette,
    breaks = c(
      "S",
      "N"
    )
  ) +
  scale_y_continuous(
    limits = y_limits,
    expand = y_expansion
  ) +
  scale_x_discrete(labels = climate_period_labels) +
  labs(
    x = "Climate Period",
    y = "Anomaly (°C)",
    fill = "Hemisphere"
  ) +
  theme_climate +
  theme(
    axis.title.y = element_blank(),
  )
```

```{r}
#| label: v3-create-composition
v3_plot_composition <-
  plt_hemisphere_avg +
  plt_hemisphere_period +
  plot_layout(
    widths = c(1, 3),
    guides = "collect"
  ) +
  plot_annotation(
    title = "Temperature Anomalies by Hemisphere",
    subtitle = "By Hemisphere and Climate Period",
    theme = theme_climate
  )
```

```{r}
#| label: v3-display-summary
#| echo: false
v3_summary
```

```{r}
#| label: v3-display-composition
#| fig-width: 12
#| fig-height: 6.5
#| echo: false
v3_plot_composition
```

### Latitude zones

Latitude-zone heatmaps examine how temperature anomalies vary geographically from the tropics to the polar regions.

```{r}
#| label: v4-prepare-heatmaps
df_zone_3_plot <- df_zone |>
  filter(
    zonal_id == "z3"
  ) |>
  mutate(
    zone = factor(
      as.character(zone),
      levels = rev(zone_levels_3)
    ),
    text_color = if_else(
      abs(avg_anomaly) >= 0.6,
      "white",
      "black"
    )
  )

df_zone_8_plot <- df_zone |>
  filter(
    zonal_id == "z8"
  ) |>
  mutate(
    zone = factor(
      as.character(zone),
      levels = rev(zone_levels_8)
    ),
    text_color = if_else(
      abs(avg_anomaly) >= 0.6,
      "white",
      "black"
    )
  )
```

```{r}
#| label: v4-color-limits
zone_color_limits <- range(
  df_zone$avg_anomaly,
  na.rm = TRUE
)
```

```{r}
#| label: v4-create-three-zone
plt_zone_3 <- df_zone_3_plot |>
  ggplot(
    aes(
      x = climate_period,
      y = zone,
      fill = avg_anomaly
    )
  ) +
  geom_tile(
    color = "white",
    linewidth = 0.5
  ) +
  geom_text(
    aes(
      label = sprintf(
        "%.2f",
        avg_anomaly
      ),
      color = text_color
    ),
    size = 6
  ) +
  scale_color_identity() +
  scale_x_discrete(
    labels = climate_period_labels
  ) +
  scale_y_discrete(
    labels = zone_labels_3
  ) +
  anomaly_fill_scale(
    zone_color_limits
  ) +
  labs(
    subtitle = "Three Latitude Zones",
    x = "Climate Period",
    y = "Latitude Zone"
  ) +
  theme_climate
```

```{r}
#| label: v4-create-eight-zone
plt_zone_8 <- df_zone_8_plot |>
  ggplot(
    aes(
      x = climate_period,
      y = zone,
      fill = avg_anomaly
    )
  ) +
  geom_tile(
    color = "white",
    linewidth = 0.5
  ) +
  geom_text(
    aes(
      label = sprintf(
        "%.2f",
        avg_anomaly
      ),
      color = text_color
    ),
    size = 6
  ) +
  scale_color_identity() +
  scale_x_discrete(
    labels = climate_period_labels
  ) +
  scale_y_discrete(
    labels = zone_labels_8
  ) +
  anomaly_fill_scale(
    zone_color_limits
  ) +
  labs(
    subtitle = "Eight Latitude Zones",
    x = "Climate Period",
    y = "Latitude Zone"
  ) +
  theme_climate
```

```{r}
#| label: v4-create-composition
v4_plot_composition <-
  plt_zone_3 / # / is used to stack vertically
  plt_zone_8 +
  plot_layout(
    heights = c(1, 2),
    guides = "collect"
  ) +
  plot_annotation(
    title = "Temperature Anomalies by Latitude Zone",
    subtitle = "Average temperature anomaly by latitude and climate period",
    theme = theme_climate
  )
```

```{r}
#| label: v4-display-composition
#| fig-width: 14
#| fig-height: 12
#| echo: false
v4_plot_composition
```

### Seasonality

Seasonal comparisons evaluate whether temperature anomalies differ systematically across the four meteorological seasons.

```{r}
#| label: v5-df-season-period
df_season_period <- df_seasonal |>
  group_by(
    season,
    climate_period
  ) |>
  summarize(
    avg_anomaly = mean(
      anomaly,
      na.rm = TRUE
    ),
    .groups = "drop"
  )
```

```{r}
#| label: v5-create-statistics
v5_statistics <- df_season_period |>
  mutate(
    season = factor(
      season,
      levels = c(
        "Winter",
        "Spring",
        "Summer",
        "Autumn"
      )
    ),
    climate_period =
      climate_period_labels[
        as.character(climate_period)
      ]
  ) |>
  pivot_wider(
    names_from = climate_period,
    values_from = avg_anomaly
  )
```

```{r}
#| label: v5-create-table
v5_summary <- v5_statistics |>
  gt() |>
  cols_label(
    season = "Season"
  ) |>
  fmt_number(
    columns = -season,
    decimals = 2
  ) |>
  cols_align(
    align = "left",
    columns = season
  ) |>
  cols_align(
    align = "center",
    columns = -season
  ) |>
  tab_header(
    title = "Average Seasonal Temperature Anomaly"
  ) |>
  tab_source_note(
    source_note =
      "Values are average temperature anomalies (°C) relative to the NASA GISTEMP 1951–1980 baseline."
  )
```

```{r}
#| label: v5-create-plot-data
df_season_period <- df_seasonal |>
  group_by(
    season,
    climate_period
  ) |>
  summarize(
    avg_anomaly = mean(
      anomaly,
      na.rm = TRUE
    ),
    .groups = "drop"
  )
```

```{r}
#| label: v5-seasonal-plot
plt_season <- df_season_period |>
  ggplot(
    aes(
      x = climate_period,
      y = avg_anomaly,
      fill = season
    )
  ) +
  geom_hline(
    yintercept = 0,
    color = "grey35",
    linewidth = 0.6
  ) +
  geom_col(
    position = position_dodge(
      width = 0.8
    ),
    width = 0.7
  ) +
  scale_fill_manual(
    values = season_palette,
    name = "Season",
    guide = guide_legend(
      direction = "horizontal",
      title.position = "top",
      title.hjust = 0.5,
      nrow = 1
    )
  ) +
  scale_x_discrete(
    labels = climate_period_labels
  ) +
  labs(
    x = "Climate Period",
    y = "Anomaly (°C)"
  ) +
  theme_climate +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    panel.grid.major.x = element_blank()
  )
```

```{r}
#| label: v5-create-composition
#| fig-width: 12
#| fig-height: 7
v5_plot_composition <-
  plt_season +
  plot_annotation(
    title = "Seasonal Temperature Anomalies",
    subtitle = "Average anomaly by season and climate period ",
    theme = theme_climate
  )
```

```{r}
#| label: v5-display-summary
#| echo: false
v5_summary
```

```{r}
#| label: v5-display-composition
#| fig-width: 12
#| fig-height: 6.5
#| echo: false
v5_plot_composition
```

### Area

Surface-area distributions quantify how much of the observed global grid experiences temperature anomalies at or above different thresholds.

```{r}
#| label: v6-weighted-quantile
weighted_quantile <- function(x, w, probs) {
  keep <- !is.na(x) & !is.na(w)
  x <- x[keep]
  w <- w[keep]
  ord <- order(x)
  x <- x[ord]
  w <- w[ord]
  cumulative_weight <- cumsum(w) / sum(w)
  sapply(
    probs,
    function(p) {
      x[which(cumulative_weight >= p)[1]]
    }
  )
}
```

```{r}
#| label: v6-create-area-exceedance
df_area_exceedance <- sf_gridded |>
  st_drop_geometry() |>
  filter(
    !is.na(avg_anomaly),
    !is.na(area_km2)
  ) |>
  arrange(
    climate_period,
    avg_anomaly
  ) |>
  group_by(climate_period) |>
  mutate(
    total_area_km2 = sum(
      area_km2,
      na.rm = TRUE
    ),
    area_at_or_above_km2 =
      rev(cumsum(rev(area_km2))),
    percent_area =
      100 *
        area_at_or_above_km2 /
        total_area_km2
  ) |>
  ungroup()
```

```{r}
#| label: v6-create-plot
#| fig-width: 12
#| fig-height: 7
plt_area_exceedance <- df_area_exceedance |>
  ggplot(
    aes(
      x = avg_anomaly,
      y = percent_area,
      color = climate_period
    )
  ) +
  geom_line(
    linewidth = 1.2,
    alpha = 0.9
  ) +
  geom_hline(
    yintercept = 50,
    color = "grey70",
    linewidth = 0.4,
    linetype = "dashed"
  ) +
  scale_color_manual(
    values = climate_period_palette,
    labels = climate_period_labels,
    name = "Climate Period"
  ) +
  scale_y_continuous(
    breaks = seq(0, 100, 20),
    labels = scales::label_percent(
      scale = 1
    )
  ) +
  coord_cartesian(
    ylim = c(0, 100)
  ) +
  labs(
    x = "Temperature Anomaly (°C)",
    y = "Observed Surface Area at or Above Anomaly"
  ) +
  theme_climate +
  theme(
    legend.position = "bottom"
  )
```

```{r}
#| label: v6-create-statistics
v6_statistics <- sf_gridded |>
  st_drop_geometry() |>
  group_by(climate_period) |>
  summarize(
    Minimum = min(
      avg_anomaly,
      na.rm = TRUE
    ),
    `25th percentile` = weighted_quantile(
      avg_anomaly,
      area_km2,
      0.25
    ),
    Median = weighted_quantile(
      avg_anomaly,
      area_km2,
      0.50
    ),
    `75th percentile` = weighted_quantile(
      avg_anomaly,
      area_km2,
      0.75
    ),
    Maximum = max(
      avg_anomaly,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  mutate(
    climate_period =
      climate_period_labels[
        as.character(climate_period)
      ]
  ) |>
  pivot_longer(
    cols = -climate_period,
    names_to = "Statistic",
    values_to = "Value"
  ) |>
  pivot_wider(
    names_from = climate_period,
    values_from = Value
  )

v6_statistics <- v6_statistics |>
  mutate(
    Statistic = recode(
      Statistic,
      `25th percentile` = "25th percentile (area-weighted)",
      Median = "Median (area-weighted)",
      `75th percentile` = "75th percentile (area-weighted)"
    )
  )
```

```{r}
#| label: v6-create-table
v6_summary <- v6_statistics |>
  gt() |>
  fmt_number(
    columns = -Statistic,
    decimals = 2
  ) |>
  cols_align(
    align = "left",
    columns = Statistic
  ) |>
  cols_align(
    align = "center",
    columns = -Statistic
  ) |>
  tab_header(
    title = "Area-Weighted Temperature Anomaly Distribution"
  ) |>
  tab_source_note(
    source_note =
      "Percentiles are weighted by observed grid-cell surface area. Temperature anomalies are relative to the NASA GISTEMP 1951–1980 baseline."
  )
```

```{r}
#| label: v6-create-composition
v6_plot_composition <-
  plt_area_exceedance +
  plot_annotation(
    title = "Global Surface Area by Temperature Anomaly",
    subtitle = "Percentage of observed surface area at or above each anomaly threshold",
    theme = theme_climate
  )
```

```{r}
#| label: v6-display-summary
#| echo: false
v6_summary
```

```{r}
#| label: v6-display-composition
#| fig-width: 12
#| fig-height: 7
#| echo: false
v6_plot_composition
```

### Global Maps

Global maps show the spatial distribution of average temperature anomalies during the REF, CSN, and REC climate periods.

```{r}
#| label: v7-map-limits
map_limits <- range(sf_gridded$avg_anomaly, na.rm = TRUE)
```

```{r}
#| label: v7-create-statistics
v7_statistics <- sf_gridded |>
  st_drop_geometry() |>
  group_by(climate_period) |>
  summarize(
    area_weighted_mean =
      weighted.mean(
        avg_anomaly,
        w = area_km2,
        na.rm = TRUE
      ),
    median_anomaly =
      median(
        avg_anomaly,
        na.rm = TRUE
      ),
    pct_area_above_0 =
      100 *
        sum(
          area_km2[avg_anomaly > 0],
          na.rm = TRUE
        ) /
        sum(
          area_km2[!is.na(avg_anomaly)],
          na.rm = TRUE
        ),
    pct_area_above_1 =
      100 *
        sum(
          area_km2[avg_anomaly > 1],
          na.rm = TRUE
        ) /
        sum(
          area_km2[!is.na(avg_anomaly)],
          na.rm = TRUE
        ),
    .groups = "drop"
  )

v7_statistics_table <- v7_statistics |>
  pivot_longer(
    cols = -climate_period,
    names_to = "statistic",
    values_to = "value"
  ) |>
  mutate(
    statistic = recode(
      statistic,
      area_weighted_mean = "Area-weighted mean anomaly",
      median_anomaly = "Median cell anomaly",
      pct_area_above_0 = "Observed area above 0 °C",
      pct_area_above_1 = "Observed area above 1 °C"
    ),
    climate_period =
      climate_period_labels[as.character(climate_period)]
  ) |>
  pivot_wider(
    names_from = climate_period,
    values_from = value
  )
```

```{r}
#| label: v7-gt-table
v7_summary <- v7_statistics_table |>
  gt() |>
  cols_label(statistic = "Statistic") |>
  fmt_number(
    rows = statistic %in% c(
      "Area-weighted mean anomaly",
      "Median cell anomaly"
    ),
    columns = -statistic,
    decimals = 2
  ) |>
  fmt_percent(
    rows = statistic %in% c(
      "Observed area above 0 °C",
      "Observed area above 1 °C"
    ),
    columns = -statistic,
    decimals = 0,
    scale_values = FALSE
  ) |>
  cols_align(
    align = "left",
    columns = statistic
  ) |>
  cols_align(
    align = "center",
    columns = -statistic
  ) |>
  tab_header(
    title = "Global Spatial Summary by Climate Period"
  ) |>
  tab_source_note(
    source_note =
      "Temperature anomalies are in °C; percentages are based on observed grid-cell area."
  )
```

```{r}
#| label: v7-display-table
#| echo: false
v7_summary
```

```{r}
#| label: v7-build-maps
if (build_maps == TRUE) {
  plt_gridded_periods <-
    sf_gridded |>
    ggplot() +
    geom_sf(
      aes(fill = avg_anomaly),
      color = NA
    ) +
    geom_sf(
      data = sf_coastline,
      inherit.aes = FALSE,
      color = "black",
      linewidth = 0.1
    ) +
    facet_wrap(
      ~climate_period,
      ncol = 1,
      labeller = as_labeller(climate_period_labels)
    ) +
    coord_sf(
      crs = mollweide_crs,
      datum = NA
    ) +
    anomaly_fill_scale(map_limits) +
    theme_void() +
    theme_climate +
    theme(
      strip.text = element_text(
        size = 15,
        face = "bold"
      )
    )
}
```

```{r}
#| label: v7-composition
if (build_maps == TRUE) {
  v7_composition <-
    plt_gridded_periods +
    plot_annotation(
      title = "Global Temperature Anomalies by Climate Period",
      subtitle = "Mean monthly temperature anomaly relative to the 1951–1980 baseline",
      theme = theme_climate
    )
}
```

```{r}
#| label: v7-display-composition
#| fig-width: 14
#| fig-height: 24
#| echo: false

v7_composition
```
